---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# climateTR

<!-- badges: start -->
![Build Status](https://img.shields.io/github/workflow/status/FrancisMsangi/climateTR/CI)
![Test Coverage](https://img.shields.io/codecov/c/github/FrancisMsangi/climateTR)
![License](https://img.shields.io/badge/license-MIT-blue.svg)
![Version](https://img.shields.io/github/v/release/FrancisMsangi/climateTR?logo=github)
![Downloads](https://img.shields.io/github/downloads/FrancisMsangi/climateTR/total)
<!-- badges: end -->

The goal of climateTR is to facilitate temporal aggregation and trend analysis of raster time-series climate data. It allows users to aggregate climate data at different temporal scales (daily, monthly, annual, and seasonal) and perform trend analysis using both parametric (linear regression) and non-parametric (Mann-Kendall) methods

## Installation

You can install the development version of climateTR from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("FrancisMsangi/climateTR")
```

## Example Usage



### Temporal Aggregation

```{r example}
library(climateTR)
library(terra)

# Generate example raster data
set.seed(123)
r_day <- rast(nrows = 10, ncols = 10, nlyrs = 365)
values(r_day) <- runif(ncell(r_day) * nlyr(r_day)) * 100

# Aggregate daily data to monthly
r_monthly <- aggr(r_day, agg_type = "monthly", temp = "day",
                  start = "2001-01-01", end = "2001-12-31", func = sum)

# Print summary
r_monthly
```

### Trend Analysis
```{r}
# Generate example raster data for trend analysis
r_year <- rast(nrows = 10, ncols = 10, nlyrs = 30)
values(r_year) <- runif(ncell(r_year) * nlyr(r_year)) * 200

# Apply linear regression trend analysis
trend_reg <- trend_rast(stacked = r_year, method = "regression")

# Apply Mann-Kendall trend analysis
trend_mann <- trend_rast(stacked = r_year, method = "Mann_Kendall")

# Print results
trend_reg
trend_mann

```

You can view the package vignette [here](https://github.com/FrancisMsangi/climateTR/blob/main/doc/ClimateTR_package_for_temporal_aggregation_and_trend_analysis_of_climate_data.html)


